---
title: "Warm-up mini-Report: Mosquito Blood Hosts in \n Salt Lake City, Utah"
author: "Ryan Colqui"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# ABSTRACT
In this study, we explored which bird species mosquitoes most commonly feed on and how that relates to West Nile Virus (WNV) activity in Salt Lake City, Utah. Using DNA extraction, PCR, and sequencing, we identified bird species from mosquito blood meals and compared sites with and without WNV positive mosquito pools. Our bar plots showed that the House Finch, House Sparrow, and American Robin were the most common hosts, with the House Finch standing out in WNV positive areas. Statistical analyses showed a strong positive relationship between the number of House Finch blood meals and WNV presence and positive rate. These results support our hypothesis that the common House Finch acts as an amplifying host for WNV in this region.

# BACKGROUND
West Nile Virus (WNV) is a disease spread mainly by mosquitoes that pick up the virus from infected birds and pass it on to other animals, including humans. Birds are really important in this cycle because they serve as the main hosts that help the virus multiply. When a mosquito bites an infected bird, it can carry the virus and spread it to new locations. Figuring out which birds mosquitoes feed on the most can help us understand which species might be acting as amplifying hosts, meaning they play a big role in spreading WNV in an area.
To find out what birds mosquitoes feed on. We looked at the DNA inside their blood meals. Scientists extract the blood from the mosquito’s stomach, then use PCR (polymerase chain reaction) to copy and amplify that DNA. After that, the DNA is sequenced and compared to known bird species so we can identify the host. In this experiment, we used this method to compare mosquito blood meals from sites with and without WNV positive mosquito pools. This helped us test our hypothesis that the common House Finch acts as an amplifying host for WNV in Salt Lake City, Utah.

## Detectable Viremia in Bird Species

```{r viremia, include=TRUE, eval=TRUE, warning=FALSE, fig.align='left', fig.width=4.75, fig.height=7}
# Manually transcribe duration (mean, lo, hi) from the last table column
duration <- data.frame(
  Bird = c("Canada Goose","Mallard", 
           "American Kestrel","Northern Bobwhite",
           "Japanese Quail","Ring-necked Pheasant",
           "American Coot","Killdeer",
           "Ring-billed Gull","Mourning Dove",
           "Rock Dove","Monk Parakeet",
           "Budgerigar","Great Horned Owl",
           "Northern Flicker","Blue Jay",
           "Black-billed Magpie","American Crow",
           "Fish Crow","American Robin",
           "European Starling","Red-winged Blackbird",
           "Common Grackle","House Finch","House Sparrow"),
  mean = c(4.0,4.0,4.5,4.0,1.3,3.7,4.0,4.5,5.5,3.7,3.2,2.7,1.7,6.0,4.0,
           4.0,5.0,3.8,5.0,4.5,3.2,3.0,3.3,6.0,4.5),
  lo   = c(3,4,4,3,0,3,4,4,4,3,3,1,0,6,3,
           3,5,3,4,4,3,3,3,5,2),
  hi   = c(5,4,5,5,4,4,4,5,7,4,4,4,4,6,5,
           5,5,5,7,5,4,3,4,7,6)
)

# Choose some colors
cols <- c(rainbow(30)[c(10:29,1:5)])  # rainbow colors

# horizontal barplot
par(mar=c(5,12,2,2))  # wider left margin for names
bp <- barplot(duration$mean, horiz=TRUE, names.arg=duration$Bird,
              las=1, col=cols, xlab="Days of detectable viremia", xlim=c(0,7))

# add error bars
arrows(duration$lo, bp, duration$hi, bp,
       angle=90, code=3, length=0.05, col="black", xpd=TRUE)
```

# STUDY QUESTION and HYPOTHESIS

## Question 
What species of bird acts as the WNV amplifying host in Salt Lake City, Utah?

## Hypothesis 
The common House Finch acts as the amplifying host of WNV in Salt Lake City, Utah.

## Prediction
If the House finch is the amplifying host for WNV, we can predict that the trapping locations which is where mosquitoes feed on the house finches will have higher rates of WNV.

# METHODS

First, We gathered mosquitos from Salt Lake City, Utah and the area surrounding it. 
Second, We then placed our mosquitos into a small tube so we are able to extract the blood's DNA from Mosquito last meal.
Third, We use our DNA we had gathered to undergo a PCR to amplify our DNA results
After we collected our amplified DNA, We put it throught the MinIon flowcell so we are able to sequence our DNA
After we got our DNA sequenced used the database of BLAST to be able to find out the animal that the mosquito had fed on.
We got results such as thel; House Sparrow, House Finch, Red-legged Seriema, Humans, and Cow just to name a few.

Horizontal Plot of what species the collected mosquitos had blood from
```{r horiz-plot,fig.align='left',fig.width=7,fig.height=7, warning=FALSE}
## import counts_matrix: data.frame with column 'loc_positives' (0/1) and host columns 'host_*'
counts_matrix <- read.csv("./bloodmeal_plusWNV_for_BIOL3070.csv")

## 1) Identify host columns
host_cols <- grep("^host_", names(counts_matrix), value = TRUE)

if (length(host_cols) == 0) {
  stop("No columns matching '^host_' were found in counts_matrix.")
}

## 2) Ensure loc_positives is present and has both levels 0 and 1 where possible
counts_matrix$loc_positives <- factor(counts_matrix$loc_positives, levels = c(0, 1))

## 3) Aggregate host counts by loc_positives
agg <- stats::aggregate(
  counts_matrix[, host_cols, drop = FALSE],
  by = list(loc_positives = counts_matrix$loc_positives),
  FUN = function(x) sum(as.numeric(x), na.rm = TRUE)
)

## make sure both rows exist; if one is missing, add a zero row
need_levels <- setdiff(levels(counts_matrix$loc_positives), as.character(agg$loc_positives))
if (length(need_levels)) {
  zero_row <- as.list(rep(0, length(host_cols)))
  names(zero_row) <- host_cols
  for (lv in need_levels) {
    agg <- rbind(agg, c(lv, zero_row))
  }
  ## restore proper type
  agg$loc_positives <- factor(agg$loc_positives, levels = c("0","1"))
  ## coerce numeric host cols (they may have become character after rbind)
  for (hc in host_cols) agg[[hc]] <- as.numeric(agg[[hc]])
  agg <- agg[order(agg$loc_positives), , drop = FALSE]
}

## 4) Decide species order (overall abundance, descending)
overall <- colSums(agg[, host_cols, drop = FALSE], na.rm = TRUE)
host_order <- names(sort(overall, decreasing = TRUE))
species_labels <- rev(sub("^host_", "", host_order))  # nicer labels

## 5) Build count vectors for each panel in the SAME order
counts0 <- rev(as.numeric(agg[agg$loc_positives == 0, host_order, drop = TRUE]))
counts1 <- rev(as.numeric(agg[agg$loc_positives == 1, host_order, drop = TRUE]))

## 6) Colors: reuse your existing 'cols' if it exists and is long enough; otherwise generate
if (exists("cols") && length(cols) >= length(host_order)) {
  species_colors <- setNames(cols[seq_along(host_order)], species_labels)
} else {
  species_colors <- setNames(rainbow(length(host_order) + 10)[seq_along(host_order)], species_labels)
}

## 7) Shared x-limit for comparability
xmax <- max(c(counts0, counts1), na.rm = TRUE)
xmax <- if (is.finite(xmax)) xmax else 1
xlim_use <- c(0, xmax * 1.08)

## 8) Plot: two horizontal barplots with identical order and colors
op <- par(mfrow = c(1, 2),
          mar = c(4, 12, 3, 2),  # big left margin for species names
          xaxs = "i")           # a bit tighter axis padding

## Panel A: No WNV detected (loc_positives = 0)
barplot(height = counts0,
        names.arg = species_labels, 
        cex.names = .5,
        cex.axis = .5,
        col = rev(unname(species_colors[species_labels])),
        horiz = TRUE,
        las = 1,
        xlab = "Bloodmeal counts",
        main = "Locations WNV (-)",
        xlim = xlim_use)

## Panel B: WNV detected (loc_positives = 1)
barplot(height = counts1,
        names.arg = species_labels, 
        cex.names = .5,
        cex.axis = .5,
        col = rev(unname(species_colors[species_labels])),
        horiz = TRUE,
        las = 1,
        xlab = "Bloodmeal counts",
        main = "Locations WNV (+)",
        xlim = xlim_use)

par(op)

## Keep the colors mapping for reuse elsewhere
host_species_colors <- species_colors
```

## First Analysis
For the first analysis, we compared the number of mosquito blood meals from each host species between locations that had no WNV positive pools and locations that had one or more WNV positive pools. This helped us see which bird species mosquitoes were feeding on the most in each type of site. We used bar plots because they make it easy to visually compare the feeding patterns across species and between WNV positive and WNV negative areas. This helps us see certain host species that the mosquitos favor.
```{r first-analysis, warning=FALSE}
## import counts_matrix: data.frame with column 'loc_positives' (0/1) and host columns 'host_*'
counts_matrix <- read.csv("./bloodmeal_plusWNV_for_BIOL3070.csv")

## 1) Identify host columns
host_cols <- grep("^host_", names(counts_matrix), value = TRUE)

if (length(host_cols) == 0) {
  stop("No columns matching '^host_' were found in counts_matrix.")
}

## 2) Ensure loc_positives is present and has both levels 0 and 1 where possible
counts_matrix$loc_positives <- factor(counts_matrix$loc_positives, levels = c(0, 1))

## 3) Aggregate host counts by loc_positives
agg <- stats::aggregate(
  counts_matrix[, host_cols, drop = FALSE],
  by = list(loc_positives = counts_matrix$loc_positives),
  FUN = function(x) sum(as.numeric(x), na.rm = TRUE)
)

## make sure both rows exist; if one is missing, add a zero row
need_levels <- setdiff(levels(counts_matrix$loc_positives), as.character(agg$loc_positives))
if (length(need_levels)) {
  zero_row <- as.list(rep(0, length(host_cols)))
  names(zero_row) <- host_cols
  for (lv in need_levels) {
    agg <- rbind(agg, c(lv, zero_row))
  }
  ## restore proper type
  agg$loc_positives <- factor(agg$loc_positives, levels = c("0","1"))
  ## coerce numeric host cols (they may have become character after rbind)
  for (hc in host_cols) agg[[hc]] <- as.numeric(agg[[hc]])
  agg <- agg[order(agg$loc_positives), , drop = FALSE]
}

## 4) Decide species order (overall abundance, descending)
overall <- colSums(agg[, host_cols, drop = FALSE], na.rm = TRUE)
host_order <- names(sort(overall, decreasing = TRUE))
species_labels <- rev(sub("^host_", "", host_order))  # nicer labels

## 5) Build count vectors for each panel in the SAME order
counts0 <- rev(as.numeric(agg[agg$loc_positives == 0, host_order, drop = TRUE]))
counts1 <- rev(as.numeric(agg[agg$loc_positives == 1, host_order, drop = TRUE]))

## 6) Colors: reuse your existing 'cols' if it exists and is long enough; otherwise generate
if (exists("cols") && length(cols) >= length(host_order)) {
  species_colors <- setNames(cols[seq_along(host_order)], species_labels)
} else {
  species_colors <- setNames(rainbow(length(host_order) + 10)[seq_along(host_order)], species_labels)
}

## 7) Shared x-limit for comparability
xmax <- max(c(counts0, counts1), na.rm = TRUE)
xmax <- if (is.finite(xmax)) xmax else 1
xlim_use <- c(0, xmax * 1.08)

## 8) Plot: two horizontal barplots with identical order and colors
op <- par(mfrow = c(1, 2),
          mar = c(4, 12, 3, 2),  # big left margin for species names
          xaxs = "i")           # a bit tighter axis padding

## Panel A: No WNV detected (loc_positives = 0)
barplot(height = counts0,
        names.arg = species_labels, 
        cex.names = .5,
        cex.axis = .5,
        col = rev(unname(species_colors[species_labels])),
        horiz = TRUE,
        las = 1,
        xlab = "Bloodmeal counts",
        main = "Locations WNV (-)",
        xlim = xlim_use)

## Panel B: WNV detected (loc_positives = 1)
barplot(height = counts1,
        names.arg = species_labels, 
        cex.names = .5,
        cex.axis = .5,
        col = rev(unname(species_colors[species_labels])),
        horiz = TRUE,
        las = 1,
        xlab = "Bloodmeal counts",
        main = "Locations WNV (+)",
        xlim = xlim_use)

par(op)

## Keep the colors mapping for reuse elsewhere
host_species_colors <- species_colors
```

## Second Analysis
For the second analysis, we ran a generalized linear model to see if the number of House Finch blood meals at each site was related to WNV activity. We made two versions of the model one that tested whether the number of House Finch meals predicted if a site had any WNV positive mosquito pools, and another that tested if it predicted how high the WNV positivity rate was.
```{r}
# second-analysis-or-plot, glm with house finch alone against binary +/_
glm1 <- glm(loc_positives ~ host_House_finch,
            data = counts_matrix,
            family = binomial)
summary(glm1)

#glm with house-finch alone against positivity rate
glm2 <- glm(loc_rate ~ host_House_finch,
            data = counts_matrix)
summary(glm2)
```

# DISCUSSION
After creating the bar plots and analyzing our data, we observed that the House Finch and House Sparrow were the most common hosts in mosquito blood meals. The second set of bar plots, which compared sites with and without WNV positive mosquito pools, showed that locations with more House Finch blood meals tended to have higher WNV positivity. To confirm this pattern, we ran a statistical analysis that revealed a significant positive correlation between the number of House Finch blood meals and the WNV positivity rate. Together, these results support our hypothesis that the House Finch acts as an amplifying host for WNV, since areas where mosquitoes feed more frequently on this species also have a higher occurrence of the virus

## Interpretation: First Analysis
When we compared the number of mosquito blood meals from each host species between locations that had no WNV positive pools and locations that had one or more WNV positive pools we noticed that there were a couple of species that stood out such as the House Finch, House Sparrow and the American Robin. With this visualization we can see that the House Finch has a significant role for being a blood meal for the mosquitoes. 

## Interpretation: Second Analysis
The results showed a significant positive relationship in both cases. Sites with more House Finch blood meals were more likely to have mosquitoes that tested positive for WNV, and their positivity rates were also higher. Overall, this supports our earlier graphs and the idea that House Finches may be acting as important amplifying hosts for West Nile Virus in our study area.

# CONCLUSION
From our results, it looks like the House Finch plays an important role in spreading West Nile Virus around Salt Lake City. Both the graphs and our data analysis showed that places with more House Finch blood meals also had higher numbers of mosquitoes that tested positive for WNV. Along with the House Sparrow and American Robin. The House Finch stood out as a common host, but it had the strongest connection to WNV. Our findings support our hypothesis that the common House Finch acts as an amplifying host for West Nile Virus in this area.

# REFERENCES     

1. Komar N, Langevin S, Hinten S, Nemeth N, Edwards E, Hettler D, Davis B, Bowen R, Bunning M. Experimental infection of North American birds with the New York 1999 strain of West Nile virus. Emerg Infect Dis. 2003 Mar;9(3):311-22. https://doi.org/10.3201/eid0903.020628

2. ChatGPT. OpenAI, version Jan 2025. Used as a reference for functions such as plot() and to correct syntax errors. Accessed `r Sys.Date()`.



